---
title: "MH survey: Data prep"
subtitle:
author:
  - name: "Eliot Monaco"
    affiliation:
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_notebook
---

# Purpose

Prepare results from the KCHD 2024 Mental Health Survey for analysis.

# Set up environment and import data

```{r setup}
knitr::opts_chunk$set(
  fig.width = 9,
  fig.height = 4
)
```

```{r message=FALSE}
library(tidyverse)
library(readxl)
library(kcData)
library(setmeup)
library(sf)
```

```{r}
source("fn.R")
```

Import survey results.

```{r}
surv_raw <- read_excel("../data/1-source/KC Health_raw data_Mental Health Survey Results 2024.xls")
```

Import codebook (created for this analysis based on the questions and response options in `MentalHealth_CityWideSurvey-FINAL_06.06.2024.docx`).

```{r}
file_path <- "../data/1-source/mh-codebook.xlsx"
sh <- excel_sheets(file_path)
codebook <- list()
for (i in 1:length(sh)) {
  codebook[[sh[i]]] <- read_excel(file_path, sheet = sh[i], col_types = "text")
}
```

Load KC population data from 2023 5-year ACS.

```{r}
pop_data <- list()
pop_data$zcta <- pop_zcta_2023acs5
pop_data$city <- pop_city_2023acs5
```

Import HOI regions.

```{r}
hoi_regions <- st_read("../data/1-source/hoi_regions.shp", quiet = TRUE)
```

# Prep data

Rename variables.

```{r}
colnames(surv_raw) <- codebook$main$nm2
```

## Validate

### `id`

Validate `id`.

```{r}
any(duplicated(surv_raw$id))
```

### Demographics

Number of non-responses to race/ethnicity question.

```{r}
surv_raw |>
  filter(if_all(d1.1:d1.8, ~ is.na(.x))) |>
  nrow()
```

Number of non-responses to birth gender question.

```{r}
surv_raw |>
  filter(is.na(d2a)) |>
  nrow()
```

Number of non-responses to current gender question.

```{r}
surv_raw |>
  filter(if_all(d2b.1:d2b.6, ~ is.na(.x))) |>
  nrow()
```

Number of non-responses to age question.

```{r}
surv_raw |>
  filter(d3 == 9) |>
  nrow()
```

### Reversed scale question

```{r}
surv_raw |>
  select(id, starts_with("q13.")) |>
  bind_cols(n_unq = apply(
    surv_raw |>
      select(starts_with("q13.")),
    1,
    function(x) length(unique(x))
  )) |>
  filter(if_all(starts_with("q13."), ~ .x %in% 1:5)) |>
  filter(n_unq == 1)
```

## Codebook

Subset main sheet from codebook and filter for MH questions.

```{r}
cbq <- codebook$main |>
  filter(str_detect(nm3, "^q"))
```

Convert response values and definitions to factor.

```{r}
# Factor levels are unique values
v <- grepl("^race|^gen|^age", names(codebook))
codebook[v] <- map(
  codebook[v],
  \(x) mutate(x, across(everything(), ~ factor(.x, levels = unique(.x))))
)

# Reverse factor levels
v <- unique(cbq[which(cbq$factor == "rev"), "codesheet", drop = TRUE])
codebook[v] <- map(
  codebook[v],
  \(x) mutate(x, across(everything(), ~ factor(.x, levels = rev(.x))))
)

# Factor levels as-is
v <- unique(cbq[is.na(cbq$factor), "codesheet", drop = TRUE])
codebook[v] <- map(
  codebook[v],
  \(x) mutate(x, across(everything(), ~ factor(.x, levels = .x)))
)
```

# Check sample characteristics

## Subset & config demo/geo table

Configure the table of demographic and geographic variables. A "Multiple" value will be included for race and current gender variables and will apply when a response to either question contained multiple options.

Race/ethnicity

- The race/ethnicity question (which allows multiple options to be selected) will be broken out into `race` and `hispanic_origin` variables.
- In the `race` variable, a race category value applies when it is the only option selected, otherwise the value is changed to "Multiple".
- `race2` is added which collapses categories with small sample sizes into larger categories. "MENA" will be added to "White", and "African" will be added to "Black".
- `race3` is added which collapses all responses that are not "Black" or "White" only into "Other".
- `hispanic_origin` values are "Hispanic" (when "Hispanic" has been selected as the only option or as one of multiple options) and "Not Hispanic" (when "Hispanic" has not been selected).

Age

- The age group values "65-74" and "75+" will be collapsed into "65+".

```{r}
surv_demo <- surv_raw |>
  # Select demographic & geographic vars
  select(id, matches("^d\\d"), zip:block_lat) |>
  mutate(
    across(c(d2a, d3), ~ as.character(.x)),
    across(c(block_lon, block_lat), ~ .x / 10^6)
  ) |>
  # Consolidate race/ethnicity vars and join definitions (original)
  unite(
    "race_united",
    matches("^d1\\.\\d$"),
    sep = "",
    remove = TRUE,
    na.rm = TRUE
  ) |>
  # Original race categories
  left_join(
    codebook$race |>
      select(val, defn),
    by = c("race_united" = "val")
  ) |>
  rename(race = defn) |>
  # Collapsed race categories (1)
  left_join(
    codebook$race2 |>
      select(val, defn),
    by = c("race_united" = "val")
  ) |>
  rename(race2 = defn) |>
  # Collapsed race categories (2)
  left_join(
    codebook$race3 |>
      select(val, defn),
    by = c("race_united" = "val")
  ) |>
  rename(race3 = defn) |>
  # New Hispanic origin variable
  mutate(
    across(c(race, race2), ~ if_else(nchar(race_united) > 1, "Multiple", .x)),
    race3 = if_else(nchar(race_united) > 1, "Other", race3),
    hispanic_origin = if_else(
      grepl("4", race_united),
      "Hispanic",
      "Not Hispanic"
    )
  ) |>
  mutate(
    race = factor(race, levels = c(
      levels(codebook$race$defn),
      "Multiple"
    )),
    race2 = factor(race2, levels = c(
      levels(codebook$race2$defn),
      "Multiple"
    )),
    race3 = factor(race3, levels = levels(codebook$race3$defn)),
    hispanic_origin = factor(
      hispanic_origin,
      levels = c("Hispanic", "Not Hispanic")
    )
  ) |>
  # Join birth gender definitions
  left_join(codebook$gen1, by = c("d2a" = "val")) |>
  rename(birth_gender = defn) |>
  # Consolidate current gender vars and join definitions
  unite(
    "cg_united",
    matches("^d2b\\.\\d$"),
    sep = "",
    remove = TRUE,
    na.rm = TRUE
  ) |>
  left_join(codebook$gen2, by = c("cg_united" = "val")) |>
  rename(current_gender = defn) |>
  mutate(current_gender = if_else(
    nchar(cg_united) > 1,
    "Multiple",
    current_gender
  )) |>
  mutate(current_gender = factor(current_gender, levels = c(
    levels(codebook$gen2$defn),
    "Multiple"
  ))) |>
  # Join collapsed age group definitions
  left_join(codebook$age2, by = c("d3" = "val")) |>
  rename(age_group = defn) |>
  select(id, zip:age_group)
```

Join HOI region variable.

```{r}
surv_rgn <- surv_demo |>
  select(id, block_lon, block_lat) |>
  st_as_sf(coords = c("block_lon", "block_lat"), crs = 4326) |>
  st_join(hoi_regions[, "region"])

surv_demo <- surv_demo |>
  bind_cols(
    surv_rgn |>
      as_tibble() |>
      select(region)
  )
```

```{r}
demo_sum <- list()
```

## Geography

Compare sampling by geography to census population counts.

### ZCTA

```{r}
demo_sum$zcta <- surv_demo |>
  group_by(zip) |>
  count() |>
  ungroup() |>
  left_join(
    pop_data$zcta |>
      filter(variable == "B01001_001") |>
      select(GEOID, estimate) |>
      mutate(GEOID = as.numeric(GEOID)),
    by = c("zip" = "GEOID")
  ) |>
  mutate(pct = pct(n, estimate, digits = 3))
```

```{r}
demo_sum$zcta
```

```{r}
demo_sum$zcta |>
  mutate(zip = factor(zip)) |>
  ggplot(aes(x = zip, y = pct)) +
  geom_col() +
  guides(x = guide_axis(angle = 90))
```

Are all ZIPs represented?

### HOI region

Summarize by region.

```{r}
demo_sum$region <- surv_demo |>
  group_by(region) |>
  count()
```

```{r}
demo_sum$region
```

```{r}
ggplot() +
  geom_sf(
    data = kcData::map_city_2024,
    fill = "darkgrey",
    color = "red"
  ) +
  geom_sf(
    data = hoi_regions,
    fill = NA,
    color = "blue"
  ) +
  geom_sf(data = surv_rgn)
```

## Demographics

Compare sampling by demographic variables to census population counts.

### Sex

```{r}
demo_sum$sex <- surv_demo |>
  group_by(birth_gender) |>
  count() |>
  ungroup() |>
  left_join(
    pop_data$city |>
      filter(variable %in% c("B01001_002", "B01001_026")) |>
      select(label, estimate) |>
      mutate(label = str_remove_all(label, "Estimate|Total|[:punct:]")),
    by = c("birth_gender" = "label")
  ) |>
  mutate(pct = pct(n, estimate, digits = 3))
```

```{r fig.width=4}
demo_sum$sex |>
  filter(!is.na(estimate)) |>
  ggplot(aes(x = birth_gender, y = pct)) +
  geom_col()
```

### Age

Select population estimates for relevant ages from the ACS.

```{r}
pop_data$age <- pop_data$city |>
  filter(
    str_detect(label, paste0("(", paste(18:100, collapse = "|"), ").*years")),
    concept == "Sex by Age"
  ) |>
  select(label, estimate) |>
  mutate(label = str_remove_all(label, "Estimate|Total|[:punct:]")) |>
  mutate(
    sex = str_extract(label, "^[:alpha:]+(?=\\d)"),
    age_group = str_extract(label, "\\d.*") |>
      str_remove("\\syears.*")
  )
```

Summarize estimates for each age range.

```{r}
# pop_data$age_ranges <- list(18:34, 35:44, 45:54, 55:64, 65:74, 75:85)
pop_data$age_ranges <- list(18:34, 35:44, 45:54, 55:64, 65:85)

pop_data$age_gp <- data.frame(
  age_group = codebook$age2$defn[1:length(pop_data$age_ranges)],
  estimate = unlist(
    lapply(
      pop_data$age_ranges,
      function(x) {
        sum(
          pop_data$age |>
            filter(str_detect(age_group, paste(x, collapse = "|"))) |>
            pull(estimate)
        )
      }
    )
  )
)
```

```{r}
demo_sum$age <- surv_demo |>
  group_by(age_group) |>
  count() |>
  ungroup() |>
  left_join(
    pop_data$age_gp,
    by = "age_group"
  ) |>
  mutate(pct = pct(n, estimate, digits = 3))

demo_sum$age
```

```{r fig.width=6}
demo_sum$age |>
  filter(!is.na(estimate)) |>
  ggplot(aes(x = age_group, y = pct)) +
  geom_col()
```

### Race

```{r}
demo_sum$race <- surv_demo |>
  group_by(race) |>
  count() |>
  ungroup()

demo_sum$race
```

```{r}
demo_sum$race2 <- surv_demo |>
  group_by(race2) |>
  count() |>
  ungroup()

demo_sum$race2
```

```{r}
demo_sum$race3 <- surv_demo |>
  group_by(race3) |>
  count() |>
  ungroup()

demo_sum$race3
```

### Hispanic origin

```{r}
demo_sum$hispanic <- surv_demo |>
  group_by(hispanic_origin) |>
  count() |>
  ungroup()

demo_sum$hispanic
```

# Subset MH questions

Subset question data into separate dataframes.

```{r}
nm <- unique(cbq$nm3)

mh <- map(
  nm,
  \(x) surv_demo |>
    bind_cols(
      surv_raw |>
        select(all_of(cbq[which(cbq$nm3 == x), "nm2", drop = TRUE]))
    )
)
names(mh) <- nm
```

# Score questions

## Q8: Pre/post COVID-19

```{r}
mh$q8 <- mh$q8 |>
  mutate(q8 = case_when(
    q8 == 4 ~ 5,
    q8 == 2 ~ 1,
    .default = q8
  ))

codebook$mh8 <- tibble(
  val = c(1, 3, 5),
  defn = c("Better", "About the same", "Worse")
) |>
  mutate(across(everything(), ~ factor(.x, levels = rev(.x))))
```

```{r include=FALSE}
# df <- data.frame(
#   q8_orig = surv_raw$q8,
#   q8_new = mh$q8$q8
# )
```

## Q11: MH treatment

Q11 should only be completed if the answer to Q10 is `1` ("Yes").

All responses to Q10.

```{r}
surv_raw |>
  mutate(val = as.character(q10)) |>
  group_by(val) |>
  count() |>
  left_join(codebook$mh9, by = "val") |>
  relocate(defn, .after = val)
```

Any rows where Q10 = `1` and any Q11 part = `NA`?

```{r}
surv_raw |>
  filter(q10 == 1) |>
  filter(if_any(q11.01:q11.10, ~ is.na(.x))) |>
  nrow()
```

Any rows where Q10 = `2` and any Q11 part = not `NA`?

```{r}
surv_raw |>
  filter(q10 == 2) |>
  filter(if_any(q11.01:q11.10, ~ !is.na(.x))) |>
  nrow()
```

Filter Q11 responses where the Q10 response is `1`.

```{r}
ids <- surv_raw |>
  filter(q10 == 1) |>
  pull(id)

mh[grepl("^q11\\.", names(mh))] <- map(
  mh[grepl("^q11\\.", names(mh))],
  \(x) x |>
    filter(id %in% ids)
)
```

## Q13: Social isolation

### Check NAs

Check for `NA` values.

```{r}
apply(
  mh$q13 |>
    select(starts_with("q13")),
  2,
  function(x) sum(is.na(x))
)
```

Count "Don't know" (`9`) responses for each question.

```{r}
apply(
  mh$q13 |>
    select(starts_with("q13")),
  2,
  function(x) sum(ifelse(x == 9, TRUE, FALSE))
)
```

### Check "Don't knows"

Number of respondents without any "Don't know" responses in Q13.

```{r}
mh$q13 |>
  filter(!if_any(starts_with("q13."), ~ .x == 9)) |>
  nrow()
```

Plot the number of "Don't know" responses per individual.

```{r}
mh$q13$n_dontknow <- apply(
  mh$q13 |>
    select(starts_with("q13.")),
  1,
  function(x) sum(ifelse(x == 9, TRUE, FALSE))
)
mh$q13$n_answered <- 11 - mh$q13$n_dontknow
```

```{r fig.width=6}
mh$q13 |>
  ggplot(aes(n_dontknow)) +
  geom_histogram(binwidth = 1)
```

### Score

Filter out rows in which any part has a "Don't know" response.

```{r}
mh$q13 <- mh$q13 |>
  filter(!if_any(starts_with("q13."), ~ .x == 9))
```

Reverse scale for questions 01, 04, 07, 08, and 11 so that `1` represents low social isolation and `5` represents high social isolation.

```{r}
mh$q13 <- mh$q13 |>
  mutate(across(
    c(q13.01, q13.04, q13.07, q13.08, q13.11),
    ~ case_when(
      .x == 1 ~ 5,
      .x == 2 ~ 4,
      .x == 4 ~ 2,
      .x == 5 ~ 1,
      .default = .x
    )
  ))
```

Sum the responses to each part to calculate a total social isolation score.

```{r}
mh$q13$q13 <- rowSums(
  mh$q13 |>
    select(starts_with("q13."))
)
```

Drop individual question variables.

```{r}
mh$q13 <- mh$q13 |>
  select(-matches("^q13\\.\\d"))
```

Assign range values.

```{r}
mh$q13 <- mh$q13 |>
  mutate(q13 = case_when(
    q13 >= 11 & q13 <= 25 ~ 1, # Low
    q13 >= 26 & q13 <= 40 ~ 2, # Medium
    q13 >= 41 & q13 <= 55 ~ 3, # High
    .default = NA
  ))
```

```{r include=FALSE}
# df <- mh$q13
```

## Q14: Social media addiction

### Check NAs

Check for `NA` values.

```{r}
apply(
  mh$q14 |>
    select(starts_with("q14")),
  2,
  function(x) sum(is.na(x))
)
```

Count "Don't know" (`9`) responses for each question.

```{r}
apply(
  mh$q14 |>
    select(starts_with("q14")),
  2,
  function(x) sum(ifelse(x == 9, TRUE, FALSE))
)
```

### Check "Don't knows"

Number of respondents without any "Don't know" responses in Q14.

```{r}
mh$q14 |>
  filter(!if_any(starts_with("q14."), ~ .x == 9)) |>
  nrow()
```

Plot the number of "Don't know" responses per individual.

```{r}
mh$q14$n_dontknow <- apply(
  mh$q14 |>
    select(starts_with("q14.")),
  1,
  function(x) sum(ifelse(x == 9, TRUE, FALSE))
)
mh$q14$n_answered <- 6 - mh$q14$n_dontknow
```

```{r fig.width=6}
mh$q14 |>
  ggplot(aes(n_dontknow)) +
  geom_histogram(binwidth = 1)
```

### Score

Filter out rows in which any part has a "Don't know" response.

```{r}
mh$q14 <- mh$q14 |>
  filter(!if_any(starts_with("q14."), ~ .x == 9))
```

Sum the responses to each part to calculate a total social media addiction score.

```{r}
mh$q14$q14 <- rowSums(
  mh$q14 |>
    select(starts_with("q14."))
)
```

Drop individual question variables.

```{r}
mh$q14 <- mh$q14 |>
  select(-matches("^q14\\.\\d"))
```

Assign range values.

```{r}
mh$q14 <- mh$q14 |>
  mutate(q14 = case_when(
    q14 >= 6 & q14 <= 13 ~ 1, # Low
    q14 >= 14 & q14 <= 22 ~ 2, # Medium
    q14 >= 23 & q14 <= 30 ~ 3, # High
    .default = NA
  ))
```

```{r include=FALSE}
# df <- mh$q14
```

## Q15: Hours on social media

Assign range values.

```{r}
mh$q15 <- mh$q15 |>
  mutate(q15 = case_when(
    q15 == 0 ~ 1,
    q15 >= 1 & q15 <= 4 ~ 2,
    q15 >= 5 & q15 <= 8 ~ 3,
    q15 > 8 & q15 <= 24 ~ 4,
    .default = NA
  ))
```

## Q16: Habitual sleep efficiency

Convert time variables to HH:MM format.

```{r}
mh$q16 <- mh$q16 |>
  mutate(
    bed1 = number_to_time(q16.1),
    wake1 = number_to_time(q16.2)
  )
```

Validate times.

```{r}
mh$q16 |>
  mutate(
    val_bed = validate_time(bed1),
    val_wake = validate_time(wake1)
  ) |>
  filter(!val_bed | !val_wake) |>
  select(q16.1:val_wake)
```

Convert bed time and wake time pairs to date objects.

```{r}
ls <- times_to_dates(mh$q16$bed1, mh$q16$wake1)
names(ls) <- c("bed2", "wake2")

mh$q16 <- mh$q16 |>
  bind_cols(ls)
```

Validate times (relative to one another).

```{r}
mh$q16 |>
  filter(bed2 > wake2)
```

Calculate the number of hours spent in bed.

```{r}
mh$q16$hrs_in_bed <- as.numeric(difftime(
  mh$q16$wake2,
  mh$q16$bed2,
  "hours"
))
```

Validate `hrs_in_bed`.

```{r}
any(mh$q16$hrs_in_bed < 0, na.rm = TRUE)
```

Calculate PSQI Component 4: Habitual sleep efficiency.

```{r}
mh$q16 <- mh$q16 |>
  mutate(hse = q16.3 / hrs_in_bed * 100) |>
  mutate(q16 = score_hse(hse))
```

View percentages > 100.

```{r}
mh$q16 |>
  filter(hse > 100) |>
  select(q16.1:last_col())
```

Filter out percentages > 100.

```{r}
mh$q16 <- mh$q16 |>
  filter(hse <= 100)
```

Drop individual question variables.

```{r}
mh$q16 <- mh$q16 |>
  select(-matches("^q16\\.\\d"))
```

## Q17: Time to fall asleep

Assign range values.

```{r}
mh$q17 <- mh$q17 |>
  mutate(q17 = case_when(
    q17 > 0 & q17 <= 10 ~ 1,
    q17 >= 11 & q17 <= 30 ~ 2,
    q17 >= 31 & q17 <= 60 ~ 3,
    q17 >= 61 & q17 <= 120 ~ 4,
    q17 >= 121 & q17 <= 3600 ~ 5,
    .default = NA
  ))
```

# Summarize results

Each question is summarized with and without a grouping variable. "Don't knows" and non-response values are not included.

```{r}
cbq <- cbq |>
  distinct(nm3, .keep_all = TRUE)

disaggs <- list(
  list(name = "full", var = NULL, label = NULL),
  list(name = "sex", var = "birth_gender", label = "Birth gender"),
  list(name = "age", var = "age_group", label = "Age"),
  list(name = "race", var = "race3", label = "Race"),
  list(name = "hispanic", var = "hispanic_origin", label = "Hispanic origin"),
  list(name = "region", var = "region", label = "HOI region")
)

mh_sum <- list()

for (i in 1:length(mh)) {
  mh_sum[[cbq$nm3[i]]] <- list()
  for (j in 1:length(disaggs)) {
    mh_sum[[i]][[disaggs[[j]]$name]] <- mh[[i]] |>
      summarize_results(
        var1 = cbq$nm3[i],
        var2 = disaggs[[j]]$var,
        cb_vals = codebook[[cbq$codesheet[i]]]$val
      ) |>
      rename(val = cbq$nm3[i]) |>
      left_join(codebook[[cbq$codesheet[i]]], by = "val")
  }
}
```

# Save

```{r}
saveRDS(cbq, "../data/3-final/codebook_question_data.rds")
saveRDS(disaggs, "../data/3-final/disaggregate_data_for_summary.rds")
saveRDS(surv_demo, "../data/3-final/survey_demographic_data.rds")
saveRDS(surv_rgn, "../data/3-final/survey_region_data.rds")
saveRDS(mh, "../data/3-final/scored_results.rds")
saveRDS(mh_sum, "../data/3-final/summarized_results.rds")
```

